---
title: "University of Washington Color Palettes in R"
author: "Joel Rosiene, MD"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{University of Washington Color Palettes in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The **UWPalette** package provides comprehensive access to the University of Washington's official web color palettes for use in R data visualization and graphics. This package implements the color schemes defined in UW's brand guidelines, ensuring consistent and professional visual presentations for academic and research purposes.

### Installation

```{r eval=FALSE}
# Install from GitHub
devtools::install_github("your-username/UWPalette")
```

```{r}
# Load the package
library(UWPalette)
```

## Available Color Palettes

The UWPalette package includes four carefully curated palettes based on University of Washington's official brand guidelines:

### Overview of All Palettes

```{r palette-overview}
# Display basic information about all palettes
palette_info <- uw_palette_info()
print(palette_info)
```

### Visual Overview

```{r palette-display, fig.height=6}
# Display all palettes visually
display_uw_palettes()
```

## Palette Details

### Primary Palette

The primary palette contains the core UW brand colors, featuring the iconic UW Purple as the primary brand color.

```{r primary-palette}
# Get primary colors
primary_colors <- uw_palette("primary")
print(primary_colors)

# Display primary palette
display_uw_palettes("primary")
```

### Secondary Palette

The secondary palette includes extended colors with metallics and neutrals that support the primary palette.

```{r secondary-palette}
# Get secondary colors
secondary_colors <- uw_palette("secondary")
print(secondary_colors)

# Display secondary palette
display_uw_palettes("secondary")
```

### Accent Palette

The accent palette provides vibrant colors for highlighting and creating visual interest.

```{r accent-palette}
# Get accent colors
accent_colors <- uw_palette("accent")
print(accent_colors)

# Display accent palette
display_uw_palettes("accent")
```

### Muted Palette

The muted palette offers subdued tones perfect for backgrounds and subtle design elements.

```{r muted-palette}
# Get muted colors
muted_colors <- uw_palette("muted")
print(muted_colors)

# Display muted palette
display_uw_palettes("muted")
```

## Basic Usage

### Getting Colors

```{r basic-usage}
# Get all colors from a palette
all_primary <- uw_palette("primary")
all_secondary <- uw_palette("secondary")

# Get specific number of colors
three_accent <- uw_palette("accent", 3)
four_muted <- uw_palette("muted", 4)

print(paste("Primary palette:", paste(all_primary, collapse = ", ")))
print(paste("First 3 accent colors:", paste(three_accent, collapse = ", ")))
```

### Palette Information

```{r palette-info}
# Get detailed information about a specific palette
accent_info <- uw_palette_info("accent")
print(accent_info)

# Get summary statistics
all_info <- uw_palette_info()
palette_summary <- aggregate(position ~ palette, data = all_info, 
                           FUN = function(x) length(x))
colnames(palette_summary) <- c("Palette", "Number_of_Colors")
print(palette_summary)
```

## Data Visualization Examples

### Base R Graphics

#### Bar Charts

```{r base-barplot}
# Create sample data
categories <- c("Category A", "Category B", "Category C", "Category D", "Category E")
values <- c(23, 45, 56, 78, 32)

# Bar plot with UW accent colors
barplot(values, 
        names.arg = categories,
        col = uw_palette("accent", 5),
        main = "Sample Bar Chart with UW Accent Palette",
        ylab = "Values",
        las = 2)
```

#### Pie Charts

```{r base-pie}
# Pie chart with muted colors
pie_data <- c(30, 25, 20, 15, 10)
pie_labels <- c("Research", "Teaching", "Service", "Administration", "Other")

pie(pie_data, 
    labels = pie_labels,
    col = uw_palette("muted", 5),
    main = "Academic Time Allocation\n(UW Muted Palette)")
```

#### Scatter Plots

```{r base-scatter}
# Scatter plot with different colors for groups
set.seed(123)
n <- 50
group1 <- data.frame(x = rnorm(n, 2, 1), y = rnorm(n, 2, 1), group = "Group 1")
group2 <- data.frame(x = rnorm(n, 4, 1), y = rnorm(n, 4, 1), group = "Group 2")
group3 <- data.frame(x = rnorm(n, 3, 1), y = rnorm(n, 6, 1), group = "Group 3")
data <- rbind(group1, group2, group3)

# Create the plot
colors <- uw_palette("secondary", 3)
plot(data$x, data$y, 
     col = colors[as.factor(data$group)],
     pch = 19,
     main = "Sample Scatter Plot with UW Secondary Colors",
     xlab = "X Variable",
     ylab = "Y Variable")
legend("topright", legend = unique(data$group), 
       col = colors, pch = 19, title = "Groups")
```

### Advanced Graphics

#### Heatmaps with Gradients

```{r heatmap}
# Create sample correlation matrix
set.seed(123)
data_matrix <- matrix(rnorm(100), nrow = 10)
correlation_matrix <- cor(data_matrix)

# Create gradient using UW colors
uw_grad <- uw_gradient("primary")
colors_gradient <- uw_grad(50)

# Create heatmap
heatmap(correlation_matrix,
        col = colors_gradient,
        main = "Correlation Heatmap\n(UW Primary Gradient)",
        symm = TRUE)
```

#### Multiple Gradients Comparison

```{r gradient-comparison, fig.height=7}
# Compare different gradient options
par(mfrow = c(2, 2), mar = c(3, 3, 3, 1))

palettes <- c("primary", "secondary", "accent", "muted")
for (pal in palettes) {
  grad_func <- uw_gradient(pal)
  colors <- grad_func(50)
  
  # Create a simple gradient display
  image(matrix(1:50, nrow = 1), 
        col = colors, 
        main = paste("UW", tools::toTitleCase(pal), "Gradient"),
        axes = FALSE,
        xlab = "", ylab = "")
}
par(mfrow = c(1, 1))  # Reset layout
```

## Integration with ggplot2

If you have ggplot2 installed, you can easily integrate UW colors:

```{r ggplot-examples, eval=FALSE}
library(ggplot2)

# Sample data
sample_data <- data.frame(
  category = LETTERS[1:6],
  values = c(23, 45, 56, 78, 32, 12),
  group = rep(c("Group 1", "Group 2"), each = 3)
)

# Bar plot with UW colors
ggplot(sample_data, aes(x = category, y = values, fill = category)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = uw_palette("accent")) +
  theme_minimal() +
  labs(title = "Sample Bar Chart with UW Accent Palette",
       subtitle = "Demonstrating UWPalette integration with ggplot2",
       x = "Category", y = "Values") +
  theme(legend.position = "none")

# Scatter plot with grouped data
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  scale_color_manual(values = uw_palette("secondary", 3),
                     name = "Cylinders") +
  theme_minimal() +
  labs(title = "MPG vs Weight by Cylinder Count",
       subtitle = "Using UW Secondary Palette",
       x = "Weight (1000 lbs)", y = "Miles per Gallon")

# Continuous scale with UW gradient
gradient_colors <- uw_gradient("muted")(100)
ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_tile() +
  scale_fill_gradientn(colors = gradient_colors) +
  theme_minimal() +
  labs(title = "Old Faithful Eruptions",
       subtitle = "Using UW Muted Gradient",
       x = "Waiting Time (minutes)", 
       y = "Eruption Duration (minutes)",
       fill = "Density")
```

## Custom ggplot2 Scales

You can create custom ggplot2 scale functions for easier integration:

```{r custom-scales, eval=FALSE}
# Custom discrete color scale
scale_color_uw <- function(palette = "primary", ...) {
  ggplot2::discrete_scale("colour", "uw", 
                         function(n) uw_palette(palette, n), ...)
}

# Custom discrete fill scale  
scale_fill_uw <- function(palette = "primary", ...) {
  ggplot2::discrete_scale("fill", "uw", 
                         function(n) uw_palette(palette, n), ...)
}

# Usage example
ggplot(sample_data, aes(x = category, y = values, fill = group)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_uw("accent") +
  theme_minimal() +
  labs(title = "Using Custom UW Scale Functions")
```

## Advanced Features

### Color Interpolation

```{r interpolation}
# Create smooth gradients between UW colors
primary_gradient <- uw_gradient("primary")
accent_gradient <- uw_gradient("accent")

# Generate different numbers of colors
colors_5 <- primary_gradient(5)
colors_10 <- primary_gradient(10)
colors_20 <- primary_gradient(20)

print("5 colors from primary gradient:")
print(colors_5)

print("10 colors from primary gradient:")
print(colors_10)
```

### Reversed Gradients

```{r reversed-gradients, fig.height=4}
# Compare normal and reversed gradients
par(mfrow = c(2, 1), mar = c(2, 3, 3, 1))

# Normal gradient
normal_grad <- uw_gradient("accent")
image(matrix(1:50, nrow = 1), col = normal_grad(50), 
      main = "Normal Accent Gradient", axes = FALSE)

# Reversed gradient
reversed_grad <- uw_gradient("accent", reverse = TRUE)
image(matrix(1:50, nrow = 1), col = reversed_grad(50), 
      main = "Reversed Accent Gradient", axes = FALSE)

par(mfrow = c(1, 1))  # Reset layout
```

### Combining Multiple Palettes

```{r combined-palettes}
# Create a custom palette combining colors from different UW palettes
custom_palette <- c(
  uw_palette("primary", 1),    # UW Purple
  uw_palette("accent", 2),     # Bright Gold and Forest Green
  uw_palette("muted", 2)       # Muted Gold and Olive Green
)

print("Custom combined palette:")
print(custom_palette)

# Use in visualization
barplot(1:5, col = custom_palette,
        main = "Custom Palette Combining UW Colors",
        names.arg = c("UW Purple", "Bright Gold", "Forest Green", 
                      "Muted Gold", "Olive Green"),
        las = 2, cex.names = 0.8)
```

## Best Practices

### Color Accessibility

When using UW colors, consider these accessibility guidelines:

1. **Contrast Ratios**: Ensure sufficient contrast between text and background colors
2. **Color Blindness**: Test visualizations with color blindness simulators
3. **Grayscale Conversion**: Verify that meaning is preserved in grayscale

```{r accessibility-demo}
# Example: Using high contrast combinations
high_contrast <- c(
  uw_palette("primary", 1),    # UW Purple
  uw_palette("secondary")[4],  # Black
  uw_palette("secondary")[5]   # White
)

print("High contrast color combination:")
print(high_contrast)
```

### Palette Selection Guidelines

- **Primary**: Use for main branding elements and primary visual hierarchy
- **Secondary**: Ideal for text, backgrounds, and secondary design elements  
- **Accent**: Use sparingly for highlighting important information
- **Muted**: Perfect for backgrounds and subtle design elements

### Professional Presentations

```{r professional-example}
# Example of a professional color scheme
professional_scheme <- list(
  title_color = uw_palette("primary", 1),      # UW Purple for titles
  text_color = uw_palette("secondary")[4],     # Black for body text
  accent_color = uw_palette("accent", 1),      # Bright Gold for highlights
  background_color = uw_palette("secondary")[5] # White for backgrounds
)

print("Professional color scheme:")
str(professional_scheme)
```

## Troubleshooting

### Common Issues

1. **Invalid palette names**: Check available palettes with `names(uw_palettes)`
2. **Requesting too many colors**: The function will return available colors if `n` exceeds palette size
3. **Integration with other packages**: Ensure color vectors are properly formatted

```{r troubleshooting}
# Check available palettes
print("Available palettes:")
print(names(uw_palettes))

# Check palette sizes
palette_sizes <- sapply(uw_palettes, length)
print("Palette sizes:")
print(palette_sizes)

# Safe color selection (won't exceed available colors)
safe_colors <- uw_palette("primary", 10)  # Only returns 2 colors
print(paste("Requested 10 colors from primary, got:", length(safe_colors)))
```

## Conclusion

The UWPalette package provides a comprehensive and easy-to-use interface for incorporating University of Washington's official colors into R visualizations. Whether you're creating simple charts or complex data visualizations, these brand-compliant color schemes ensure professional and consistent visual presentations.

### Key Features Summary

- **Four comprehensive palettes**: Primary, Secondary, Accent, and Muted
- **Flexible color selection**: Get any number of colors from each palette
- **Gradient support**: Create smooth color transitions for continuous data
- **Easy integration**: Works seamlessly with base R graphics and ggplot2
- **Comprehensive documentation**: Detailed help and examples for all functions

### Getting Help

For additional help and examples:

- Use `?uw_palette` for function documentation
- Use `?uw_palettes` for palette data information  
- Check the package README on GitHub
- View function examples with `example(uw_palette)`

---

*This vignette demonstrates the core functionality of the UWPalette package. For the most up-to-date information and additional examples, please visit the package GitHub repository.*